---
title: "Codebook"
author: "Luna"
date: "4/7/2020"
output: html_document: keep_md:yes
---

## Project Description
Human Activity Recognition database built from the recordings of 30 subjects performing activities of daily living (ADL) while carrying a waist-mounted smartphone with embedded inertial sensors.
Here, the project cleans the data collected from the experiments and prepares it for further analysis.  

##Study design and data processing  

###Collection of the raw data
The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

###Notes on the original (raw) data 
 
A video of the experiment including an example of the 6 recorded activities with one of the participants can be seen in the following link: [Web Link]<http://www.youtube.com/watch?v=XOEN9W05_4A>  

An updated version of this dataset can be found at [Web Link]<http://archive.ics.uci.edu/ml/datasets/Smartphone-Based+Recognition+of+Human+Activities+and+Postural+Transitions>. It includes labels of postural transitions between activities and also the full raw inertial signals instead of the ones pre-processed into windows.

##Creating the tidy datafile

###Guide to create the tidy data file
1. Download the data from the [website]<http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones>. 
2. Store it in your working directory.  
3. Run the script ("run_analysis.R").  
4. The script will automatically output two files in your working directory: 
* "TidyDataset.txt"--the tidy version of the dataset which contains only the measurements on the mean and standard deviation for each measurement.  
* "TidyAverage.txt"--a tidy data set with the average of each variable for each activity and each subject.  

###Cleaning of the data
In general, what the cleaning script does is to merge the seperated training and test datasets into one and add the corresponding activities and subjects labels that each measurements apply to.   
After that, we select only the measurements on the mean and standard deviation for each measurement in order to perform further analysis.   
Finally we create another dataset with the average of each variable for each activity and each subject.
Please check the script for more details.

##Description of the variables in the TidyDataset.csv file

General description of the file:
```{r}
# First please put the dataset in your working directory
TidyDataset<-read.csv("TidyDataset.txt",sep=" ")
dim(TidyDataset) # Dimensions of the dataset
str(TidyDataset) # Summary of the data
names(TidyDataset) # Variables present in the dataset
```

###Variable 1 : Activity
```{r}
str(TidyDataset$Activity)
table(TidyDataset$Activity)
```
The first variable Activity is a factor with 6 levels and has no units since it is a categorical variable.
It describes which of the six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) were the subject performing when recording the following measurements.

###Variable 2 : Subject
```{r}
str(TidyDataset$Subject)
table(TidyDataset$Subject)
```
The second variable Subject is a factor with 30 levels, each level represent a subject who carried out the experiment. Also it has no units.
In total, there are 30 people enrolled in the project.

###Other Variables : Measurements
```{r}
str(TidyDataset[3:81])
summary(TidyDataset[3:81])
```
Other variables included in the dataset are also called features which come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. They are all numeric vectors ranging from -1 to 1. Here is how the measurements generated and their respective units:

1. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 
2. Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 
3. Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

tBodyAcc-XYZ
tGravityAcc-XYZ
tBodyAccJerk-XYZ
tBodyGyro-XYZ
tBodyGyroJerk-XYZ
tBodyAccMag
tGravityAccMag
tBodyAccJerkMag
tBodyGyroMag
tBodyGyroJerkMag
fBodyAcc-XYZ
fBodyAccJerk-XYZ
fBodyGyro-XYZ
fBodyAccMag
fBodyAccJerkMag
fBodyGyroMag
fBodyGyroJerkMag

The set of variables that were estimated from these signals are: 

mean(): Mean value
std(): Standard deviation
mad(): Median absolute deviation 
max(): Largest value in array
min(): Smallest value in array
sma(): Signal magnitude area
energy(): Energy measure. Sum of the squares divided by the number of values. 
iqr(): Interquartile range 
entropy(): Signal entropy
arCoeff(): Autorregresion coefficients with Burg order equal to 4
correlation(): correlation coefficient between two signals
maxInds(): index of the frequency component with largest magnitude
meanFreq(): Weighted average of the frequency components to obtain a mean frequency
skewness(): skewness of the frequency domain signal 
kurtosis(): kurtosis of the frequency domain signal 
bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.
angle(): Angle between to vectors.

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

gravityMean
tBodyAccMean
tBodyAccJerkMean
tBodyGyroMean
tBodyGyroJerkMean

The complete list of variables of each feature vector is available in 'features.txt' which can be found in [the original data]<http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones>. 

##Sources
Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. A Public Domain Dataset for Human Activity Recognition Using Smartphones. 21th European Symposium on Artificial Neural Networks, Computational Intelligence and Machine Learning, ESANN 2013. Bruges, Belgium 24-26 April 2013.

A full description is available at the [site]<http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones> where the data was obtained. 
